Interpretable Neural Networks using EAGGA

### 2025-03-14 paper/main.tex
interpretability evaluation

% feels better in front of AutoML, as it feels more qualitative, less quantitative
\subsection{Interpretability}
As there is no clear definition for interpretability, we will consider it as ``the ability to provide \textit{explanations} in \textit{understandable terms} to a human'',
where \textit{explanations} are logical decision rules and \textit{understandable terms} relate to commonly used terms in the domain of the problem,
as suggested by \citet[chap. 1]{survey_NN_interpretability}. Further, we use the term ``explainability'' in an exchangeable manner with ``interpretability'',
as is commonly done.

@@

% Evaluation
- objective evaluation difficult, as no clear definition of interpretability
- \citet[3]{DoshiVelez2017TowardsAR} propose a taxonomy to categorise possible evaluation methods
  * \textbf{application-grounded evaluation}
    - most rigorous method, also most expensive + time consuming
    - idea: evaluate ``interpretability model'' directly w.r.t. the task -> human experts evaluate the outcome
    - e.g.: model performing medical diagnosis on patients would be evaluated based on doctors doing the same
  * \textbf{human-grounded metrics}
    - similar to application-grounded, but tries to simplify task (while preserving its essence) so that laypersons can do it
    - cheaper due to larger, less qualified subject pool
    - especially suitable if only general concepts of the tasks need to be validated
    - examples for evaluation set-up
      * binary forced choice: human evaluator chooses which of two explanations he finds better
      * forward simulation: evaluator must correctly simulate model output when presented model input + explanation % TODO: perhaps binary forced choice is already enough to include, this (and the third one is a bit awkward to explain why this is good for interpretability imo)
  * \textbf{functionally-grounded evaluation}
    - assess explanation quality via some formally defined proxy for interpretability
    - no human time nor cost required beyond initial formulation of proxy
    - e.g. if we already have an interpretable model class (e.g. identified via human-grounded evaluation), can then rank different models of that class based on proxy
    - main challenge: what proxies

>>>>

\subsection{Interpretability}
As there is no clear definition for interpretability, we will consider it as ``the ability to provide \textit{explanations} in \textit{understandable terms} to a human'',
where \textit{explanations} are logical decision rules and \textit{understandable terms} relate to commonly used terms in the domain of the problem,
as suggested by \citet[chap. 1]{survey_NN_interpretability}. Further, we use the term ``explainability'' in an exchangeable manner with ``interpretability'',
as is commonly done.

@@

% Evaluation
Given its unclear definition, evaluating interpretability can be challenging. \citet[3]{DoshiVelez2017TowardsAR} propose a taxonomy to categorise
possible evaluation methods based on their rigorousness.\\
\textbf{Application-grounded evaluation} evaluates interpretations directly with respect to the task, by having human experts evaluate the outcome and
is therefore the most expensive and time-consuming of the three approaches.\\
\textbf{Human-grounded metrics} is similar to application-grounded evaluation in that a human still evaluates the interpretations, but tries to simplify the task
so that a layperson can do it. Human-grounded approaches are especially suitable if it's sufficient to validate the general concepts of a task. A typical evaluation
set-up in this category is binary forced choice, where the human evaluator chooses, which of two generated explanations he prefers.\\
\textbf{Functionally-grounded evaluation} is the least rigorous, but easiest to implement of the three. It assess explanatory quality according to some
formally defined proxy for interpretability and is particularly useful in ranking different models if their model-class is already identified
(e.g. via human-grounded evaluation) to be interpretable. The main challenge for functionally-grounded evaluation is finding a good proxy.
